\textsc{HubbleCLIP}: Associating Astronomical Observations and Natural Language with Multi-Modal Models

### 2024-02-19 paper/main.tex
update paper with new results

\begin{abstract}
We present a multi-modal model which associates astronomical observations imaged by the \hubble Space Telescope (HST) with natural language. The model is fine-tuned from a base Contrastive Language–Image Pre-training (CLIP) model using successful proposal abstracts corresponding to HST observations, summarized via guided large language model (LLM) generation. We show that the model embodies a meaningful joint representation between observations and text through experiments targeting observation retrieval (i.e., retrieving the most relevant observations from a set using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases most relevant to a given observation). The model demonstrates the potential for using generalist rather than task-specific models for astrophysics research, in particular by leveraging text as an interface.
\end{abstract}

@@

Machine learning (ML) is beginning to have a significant impact in the sciences, with astrophysics being no exception. ML methods have demonstrated promise at every stage of the astrophysics pipeline, from instrument design, to data acquisition, to its analysis \citep{huertas2022dawes}. Until recently, most applications of ML within astrophysics have focused on replacing or augmenting traditional techniques with ML counterparts in order to improve performance on specific tasks. The \emph{Foundation Model} paradigm, in contrast, seeks to develop generalist models which can be deployed on a wide range of tasks. The paradigm has been highly successful in domains like computer vision and natural language processing, as demonstrated by the widespread adoption of tools like ChatGPT, Dall-E, Stable Diffusion, and CLIP. These models are typically pre-trained on massive amounts of unlabeled data using self-supervised learning techniques, enabling them to learn powerful representations which can be deployed on different tasks across domains. At the heart of the paradigm lies the triumph of scale -- scaling up model size, dataset size, and compute. However, foundation models can often benefit from \emph{fine tuning} using a small amounts of domain data, increasing their usefulness when applied to those target domains.

There is considerable interest in developing custom foundation models for the sciences \citep{batatia2023foundation,subramanian2023towards}, with astrophysics being ripe for such an effort given the availability of large amounts of publicly-available data and diverse ways of interacting with it. The multi-modality inherent to the nature of astrophysical observations, with different types of data (e.g., images, spectra, light curves, textual descriptions) often available, presents a unique opportunity. This multi-modality was recently exploited in \textsc{AstroCLIP}~\citep{lanusse2023astroclip} -- method to obtain a joint embedding space between multi-band images and optical spectra from the Dark Energy Spectroscopic Instrument (DESI). \textsc{AstroLLaMA}~\citep{nguyen2023astrollama} is another recent effort to fine-tune a publicly-available model (\textsc{Llama-2}) on astrophysics-specific textual data from the arXiv.

The Contrastive Language-Image Pre-training~\citep[CLIP;][]{radford2021learning} family of foundation models, which in its original instantiation embed images and text-via-captions into a common embedding space via contrastive learning, has shown strong performance and generalization capabilities on a variety of downstream tasks including zero-shot classification and image retrieval. Besides \textsc{AstroCLIP}, the concept of contrastive pre-training has been used in a variety of domains, e.g \textsc{GeoCLIP}~\citep{cepeda2023geoclip} for geospatial location querying.

The success of the foundation model paradigm partly relies on the ability to flexibly leverage text as a \emph{universal interface} for interacting with models. In this work, we take this outlook and train a joint representation between observations taken by the \hubble Space Telescope (HST) and natural language. We do so by leveraging associations between LLM-summarized abstracts of successful observing proposals and corresponding downstream image observations. We show that fine-tuning a pre-trained CLIP image-text model on this data enables learning semantically meaningful joint representations through quantitative and qualitative evaluation on retrieval tasks.

@@

We curate a dataset of images of \hubble observations and corresponding text descriptions. We rely on summarized versions of proposal abstracts from the Proposal Abstracts Catalog\footnote{\url{https://archive.stsci.edu/hst/proposal_abstracts.html}} -- a catalog of all accepted \hubble proposals to-date -- to derive captions for the observations. The HST has been operational for 33 years, having been launched on April 24, 1990. We use available proposals and observations up to Cycle 30, which commenced data-taking in 2022.

Examples of images and corresponding captions are shown in Tab.~\ref{tab:dataset}. It can be seen that these images have specific characteristics as well as artifacts particular to the nature of data-taking which distinguish them from the distribution of natural images typically used for large-scale pre-training of foundation models. This further motivates the need for fine-tuning on domain-specific data.

@@

Observations corresponding to individual proposal IDs are queried through the Mikulski Archive for Space Telescopes (MAST) via the \package{Astroquery} interface. Products of type \texttt{PREVIEW} are filtered in, corresponding to preview postcard images. We note that these are not science-grade observations, but rather lower-resolution images useful for quick-look purposes; given the nature of associations we aim to learn, this is adequate for our purposes. A maximum of 20 images are downloaded per proposal ID, selected at random, in order to avoid biasing the model towards proposals with a larger number of observations and survey-style campaigns. Images are centered and resized to a resolution-per-side of 512 pixels before. Color previews (i.e., observations taken with multiple wavelength filters assigned to individual RGB channels) are manually excluded via a filename filter in order to maintain consistency across the dataset; models trained on datasets with color images included were observed to show worse performance on generalization metrics. If no appropriate images corresponding to an abstract are found, is is excluded from the dataset.

@@

@@

\subsection{Summarization via Guided Generation}
\label{sec:summarization}

Raw abstracts summarize the corresponding successful HST observing proposals, which intend to make the case for allocating \hubble telescope time towards a particular set of observations. These abstracts are written in a diversity of styles, formats, and lengths, being highly variable in the nature of content as well. Although the abstracts can be used as-is as image captions, for our baseline model we summarize them via guided LLM generation to standardize the captions used for fine-tuning the CLIP model. The goal is to summarize the objects and phenomena, as well as potential downstream science use cases corresponding to the HST observations in order to increase the strength of the association signal between text and images.

@@

The procedure guides the generation of LLM outputs while ensuring that the schema is respected at every point in the generation by masking out tokens that would violate the intended format. By framing the problem in terms of transitions between a set of finite-state machines, \cite{willard2023efficient} showed that guided generation can be performed with negligible overhead compared to unconstrained generation. This ensures that the output of the LLM strictly conforms to the format of the following example:
\begin{lstlisting}[language=Python]
{
  'objects_and_phenomena': ['star forming galaxy', 'lensed galaxy', ...], 
  'science_use_cases': ['measure lensing magnification', 'probe spectral energy distributions', ...]
}
\end{lstlisting}
which is then used to construct the summarized caption by combining the two key elements. Examples of raw abstract snippets and corresponding LLM-generated summaries are shown in Tab.~\ref{tab:datasetsumm}. See App.~\ref{app:guided-generation} for a more detailed description of this guidance generation process.

@@

We emphasize that the goal of summarization-via-guided-generation is to increase the signal between text and images by standardizing the captions used for fine-tuning the CLIP model, and compare the quantitative performance of the model vs using the raw abstracts in Sec.~\ref{sec:results}. We also note that, even after summarization, the association signal is expected to be noisy, since the summarized caption may not always be usefully descriptive of the observed images.

Finally, in order to test a further compression of the proposal abstracts, we use LLM-guided generation to also produce a list of single-concept summaries (e.g., `irregular galaxy', `quasars', `Galactic bulge', \ldots), and test whether this can lead to meaningful, generalizable learned associations with observations. The prompts and schemata for generating and assigning these are described in Apps.~\ref{app:singleconcept} and \ref{app:singleconceptassignments} respectively.

@@

\begin{table}[h!]
      \centering
      \begin{tabular}{m{0.21\textwidth} p{1.9cm} p{1.9cm} m{8cm}}
          \toprule
          \centering \bfseries Image & \centering \bfseries Obs. cycle & \centering \bfseries Prop. ID & \centering \bfseries LLM-extracted summary \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_0.pdf} & \centering \input{\thedatafolder/cycle_0.txt} & \centering \input{\thedatafolder/id_0.txt} &  {\scriptsize \input{\thedatafolder/sum_0.txt}} \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_1.pdf} & \centering \input{\thedatafolder/cycle_1.txt} & \centering \input{\thedatafolder/id_1.txt} &  {\scriptsize \input{\thedatafolder/sum_1.txt}} \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_2.pdf} & \centering \input{\thedatafolder/cycle_2.txt} & \centering \input{\thedatafolder/id_2.txt} &  {\scriptsize \input{\thedatafolder/sum_2.txt}} \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_3.pdf} & \centering \input{\thedatafolder/cycle_3.txt} & \centering \input{\thedatafolder/id_3.txt} &  {\scriptsize \input{\thedatafolder/sum_3.txt}} \tabularnewline
          \bottomrule
      \end{tabular}
      \caption{Examples of images and corresponding captions, constructed using the LLM-extracted summaries. The CLIP model is fine-tuned on these text-image associations. \SM{Can add another row to fill the page}}
      \label{tab:dataset}
  \end{table}

\begin{landscape}
      \begin{table}[h!]
          \centering
          \begin{tabular}{m{1.8cm} m{8cm} m{5cm} m{6cm}}
              \toprule
              \bfseries Prop. ID & \centering\arraybackslash \bfseries Proposal abstract & \multicolumn{2}{c}{\bfseries LLM-extracted summary} \tabularnewline
              \cmidrule(r){3-4}
              & & \centering\arraybackslash \bfseries Objects and phenomena & \centering\arraybackslash \bfseries Science use cases \tabularnewline
              \midrule
              \input{\thedatafolder/id1_0.txt} & {\scriptsize \input{\thedatafolder/abs1_0.txt}} & {\scriptsize \input{\thedatafolder/obj1_0.txt}} & {\scriptsize \input{\thedatafolder/sci1_0.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_1.txt} & {\scriptsize \input{\thedatafolder/abs1_1.txt}} & {\scriptsize \input{\thedatafolder/obj1_1.txt}} & {\scriptsize \input{\thedatafolder/sci1_1.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_2.txt} & {\scriptsize \input{\thedatafolder/abs1_2.txt}} & {\scriptsize \input{\thedatafolder/obj1_2.txt}} & {\scriptsize \input{\thedatafolder/sci1_2.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_3.txt} & {\scriptsize \input{\thedatafolder/abs1_3.txt}} & {\scriptsize \input{\thedatafolder/obj1_3.txt}} & {\scriptsize \input{\thedatafolder/sci1_3.txt}} \tabularnewline
              \bottomrule
          \end{tabular}
          \caption{Examples of the initial parts of raw proposal abstracts (second column) and LLM (\textsc{Mixtral-8x7B})-extracted summaries (rightmost two columns), separately extracting objects and phenomena as well as potential downstream science use cases. The LLM-extracted summaries are used for associating text with observations.}
          \label{tab:datasetsumm}
      \end{table}
  \end{landscape}

@@

Our goal is to learn a semantically  meaningful joint representation between HST image observations and natural language. We leverage the strong generalization capabilities demonstrated by pre-trained CLIP models and adapt these to work with domain-specific \hubble data via fine-tuning.

@@

CLIP \citep[Contrastive Language-Image Pretraining;][]{radford2021learning} is a multi-modal model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss. Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$. Image and text encoders $f: I \rightarrow \mathbb R^{n_\text{emb}}$ and $g: T \rightarrow \mathbb R^{n_\text{emb}}$ are used to map images and text to a common embedding space of dimension $n_\text{emb}$. The standard softmax-based bidirectional variation of the InfoNCE~\citep{oord2018representation} contrastive loss function was introduce for training CLIP-style architectures\citep{radford2021learning}
\begin{equation}
  \mathcal{L}(\mathcal{B})=-\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_i \cdot y_j / \tau}}+\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_j \cdot y_i / \tau}}\right)\label{eq:softmax_loss}
\end{equation}
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter. Note that this loss treats the image and text representations symmetrically, ensuring that the two modalities are on the same footing.

We use the CLIP-ViT-B/16 \citep{radford2021learning} variant as the base pre-trained CLIP model. This model uses a 12-layer, 12-head, 768-embedding dimension vision transformer as the image encoder and a 12-layer, 8-head, 512-embedding dimension sequence transformer as the text backbone. The text encoder has a maximum length of 77 tokens and the image encoder a native resolution of $224\times224$. Linear projection layers map the outputs of the image and text encoders to a common embedding space of dimension 512. In total, the model has 149,620,737 trainable parameters. The model was originally trained on 400 million image-text pairs from internet data.

@@

The base CLIP model is fine-tuned using the dataset described in Sec.~\ref{sec:dataset}, using either the LLM-summarized data or the raw proposal abstracts. When using raw proposal abstracts, random chunks of the text delimited by periods are selected on the fly to fit within the maximum token length of the text encoder. Images are randomly cropped to the native resolution of the image encoder and randomly rotated at each training step. Given the relatively modest size of the fine-tuning set, a batch size $|\mathcal B| = 32$ is used throughout; larger batch sizes were seen to susceptible to overfitting. We note that the positive and negative image-text association is noisy and imperfect, since multiple images can be associated with the same abstract.

@@

Models were instantiated and trained using the \package{Jax} \citep{jax2018github} ecosystem. Models were trained over 20,000 steps with 2000 linear warmup steps 
% and cosine decay 
using the AdamW optimizer \citep{DBLP:conf/iclr/LoshchilovH19,DBLP:journals/corr/KingmaB14} with  %peak 
learning rate of $10^{-5}$ and weight decay $10^{-3}$. Training takes approximately 3 hours on 4 Nvidia A100 GPUs.

@@

The model is evaluated by tracking the loss in Eq.~\ref{eq:softmax_loss} as well as the top-$k\%$ retrieval accuracy on the held out validation set over the course of training. The retrieval accuracy is defined as the fraction of associated captions which fall within the top $k\%$ of captions by cosine similarity of the (normalized) embeddings $x_i \cdot y_j$, averaged over the images in the validation set.

@@

Figure~\ref{fig:retrieval_acc} shows the contrastive loss (left) and the top-10\% retrieval accuracy on the held out validation set over the course of training, for different variations considered. The red lines show the metrics evaluated when training with batches where the image-text associations are randomly shuffled, serving as a baseline. This baseline is seen to do on par with random expectation, unlike the others, validating the presence of a significant association signal in the dataset.
%
Interestingly, the base model performs better than random expectation, with a top-10\% retrieval accuracy of $\sim 15\%$. We therefore compare the qualitative performance of the base model with the fine-tuned model on downstream retrieval tasks.

The fiducial model with LLM-guided summarization (orange lines) is seen to perform significantly better than the model using raw abstracts as captions (purple line), validating the stronger association signal in the summarized dataset curated. Fine-tuning a small MLP head over frozen vision and text backbones (green lines) and training from scratch (blue lines) show a non-trivial improvement compared to the random baseline as well as base model, but with deteriorated performance compared to the fiducial set-up.

\begin{figure*}[!h]
  \includegraphics[width=0.95\textwidth]{plots/val_metrics.pdf}
  \caption{The CLIP-contrastive loss from Eq.~\ref{eq:softmax_loss} and the top-10\% retrieval accuracy computed on the validation set over the course of training. Shown for the fiducial set-up (orange), dataset using raw proposal abstracts (purple), only fine-tuning a small MLP head (green), training from scratch (blue), and a baseline trained with shuffled image-text pairs.}
  \label{fig:retrieval_acc}
  \end{figure*}

@@

Having aligned the image and text representations, we can query the validation set using natural language and show the most ``relevant'' images when ranked by cosine similarity. We show these in Figs.~\ref{fig:tti_base} and \ref{fig:tti} for the base model and fine-tuned model respectively using four simple curated queries: `globular clusters', `dwarf galaxy', `SN1987A' (a specific, flagship supernova), and `cluster lensing'.

The base model shows signs of meaningful retrieval -- it returns observations that clearly visually resemble globular clusters (first row of Fig.~\ref{fig:tti_base}), for example. Beyond this, the `dwarf galaxy' query returns a mix of galaxies and globular clusters, and it is challenging to discern meaningful associations between the other retrieved images and corresponding query.

>>>>

\begin{abstract}
We train a multi-modal model which associates astronomical observations imaged by the \hubble Space Telescope (HST) with natural language. The model is fine-tuned from a pre-trained Contrastive Language–Image Pre-training (CLIP) model using successful proposal abstracts corresponding to HST observations, optionally summarized via guided large language model (LLM) generation. We show that the model embodies a meaningful joint representation between observations and text through experiments targeting image retrieval (i.e., retrieving the most relevant observations from a set using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases most relevant to a given observation). Our study demonstrates the potential for using generalist rather than task-specific models for astrophysics research, in particular by leveraging text as an interface.
\end{abstract}

@@

Machine learning (ML) is beginning to have a significant impact in the sciences, with astrophysics being no exception. ML methods have demonstrated promise at every stage of the research pipeline, from instrument design, to data acquisition, to its analysis \citep{huertas2022dawes}. Until recently, most applications of ML within astrophysics have focused on replacing or augmenting traditional techniques with ML counterparts in order to improve performance on specific, individual tasks. The \emph{Foundation Model} paradigm, in contrast, seeks to develop generalist models which can be deployed to tackle a wide range of tasks. The paradigm has been highly successful in domains like computer vision and natural language processing, as demonstrated by the widespread adoption of tools like ChatGPT, Dall-E, Stable Diffusion, and CLIP. These models are typically pre-trained on massive amounts of unlabeled data using self-supervised learning techniques, enabling them to learn powerful representations which can be deployed on different tasks across domains. At the heart of the paradigm lies the triumph of scale -- scaling up model size, dataset size, and compute. However, foundation models can often benefit from additional training ({fine tuning}) using a relatively small amounts of domain-specific data in order to increase their usefulness when applied to specialized domains.

There is considerable interest in developing custom foundation models for the sciences \citep[e.g., ][]{batatia2023foundation,subramanian2023towards}, with astrophysics being ripe for such an effort given the availability of large amounts of publicly-available data and diverse ways of interacting with it. The multi-modality inherent to the nature of astrophysical observations, with different types of data (e.g., images, spectra, light curves, textual descriptions) often available for a given object, presents a unique opportunity. This multi-modality was recently exploited in \textsc{AstroCLIP}~\citep{lanusse2023astroclip} to obtain a joint embedding space between multi-band images and optical spectra from the Dark Energy Spectroscopic Instrument (DESI). \textsc{AstroLLaMA}~\citep{nguyen2023astrollama,perkowski2024astrollama} is another recent effort to fine-tune a publicly-available model (\textsc{Llama-2}) on astrophysics-specific textual data from the arXiv.

The Contrastive Language-Image Pre-training~\citep[CLIP;][]{radford2021learning} family of foundation models, which in their original instantiation embed images and associated captions into a common embedding space via contrastive learning, have shown strong performance and generalization capabilities on a variety of downstream tasks including zero-shot classification and image retrieval. The concept of associating diverse modalities via contrastive training has been used on other scientific domains~\citep[e.g.,][]{liu2023text,Sanchez-Fernandez2022.11.17.516915,lanusse2023astroclip,cepeda2023geoclip}, and has been shown to be effective in learning semantically meaningful joint representations.

The success of the foundation model paradigm relies in part on the ability to flexibly leverage text as a \emph{universal interface} for interacting with models. In this work, we take this outlook and train a joint representation between observations taken by the \hubble Space Telescope (HST) and natural language. We do so by leveraging associations between LLM-summarized abstracts of successful observing proposals and corresponding downstream image observations. We show that fine-tuning a pre-trained CLIP image-text model on this data results in meaningful joint representations through quantitative and qualitative evaluation.

@@

We curate a dataset of images of \hubble observations and corresponding text descriptions. We rely on proposal abstracts from the Proposal Abstracts Catalog\footnote{\url{https://archive.stsci.edu/hst/proposal_abstracts.html}} -- a catalog of all accepted \hubble proposals to-date -- to derive captions for the observations, optionally summarizing the abstract via guided LLM generation (described in Sec.~\ref{sec:summarization} below). The HST has been operational since its launch on April 24, 1990, and we use available proposals and observations up to Cycle 30, which commenced data-taking in 2022.

Examples of images and corresponding (summarized) captions are shown in Tab.~\ref{tab:dataset}. It can be seen that these images have specific characteristics as well as artifacts particular to the nature of data-taking and processing which distinguish them from the distribution of natural images typically used for large-scale pre-training of foundation models. This further motivates the need for fine-tuning on domain-specific data.

@@

Observations corresponding to individual proposal IDs are queried through the Mikulski Archive for Space Telescopes (MAST)\footnote{\url{https://mast.stsci.edu/}} via the \package{Astroquery} \citep{2019AJ....157...98G} interface. Products of type \texttt{PREVIEW} are filtered in, corresponding to preview postcard images. We note that these are not science-grade observations, but rather lower-resolution images useful for diagnostic purposes. A maximum of 20 images are downloaded per proposal ID, selected at random, in order to avoid biasing the model towards proposals with a larger number of observations and survey-style campaigns. Images are centered and resized to a resolution-per-side of 512 pixels before. Color previews (i.e., observations taken with multiple wavelength filters assigned to individual RGB channels) are manually excluded via a filename filter in order to maintain consistency across the dataset; models trained on datasets with color images included were observed to show worse performance on generalization metrics. If no appropriate images corresponding to an abstract are found, it is excluded from the dataset.

@@

We note that some fraction of the image-caption pairs in the constructed dataset will center instrumentation and/or calibration rather than primarily scientific content. We choose to not filter out these pairs from our dataset, in order to have a larger sample of HST observations that the image encoder can leverage to adapting to the distinctive characteristics of \hubble images.

@@

\subsection{Abstract Summarization via Guided Generation}
\label{sec:summarization}

Raw proposal abstracts summarize the corresponding successful HST observing proposals, which intend to make the case for allocating \hubble telescope time towards a particular set of observations. These abstracts are written in a diversity of styles, formats, and lengths, being highly variable in the nature of content as well. Although the abstracts can be used as-is as image captions, we experiment with summarizing them via guided large language model (LLM) generation to standardize the captions used for fine-tuning the CLIP model. The goal is to summarize the objects and phenomena, as well as potential downstream science use cases corresponding to the HST observations in order to increase the strength of the association signal between text and images.

@@

The procedure guides the generation of LLM outputs while ensuring that the schema is respected at every step in the autoregressive generation process by masking out tokens that would violate the intended format. By framing the problem in terms of transitions between a set of finite-state machines, \cite{willard2023efficient} showed that guided generation can be performed with negligible overhead compared to unconstrained generation. This ensures that the output of the LLM strictly conforms to the format of the following example:
\begin{lstlisting}[language=Python]
{
  'objects_and_phenomena': ['star forming galaxy', 'lensed galaxy', ...], 
  'science_use_cases': ['measure lensing magnification', 'probe spectral energy distributions', ...]
}
\end{lstlisting}
which is then used to construct the summarized caption by combining the two key elements. Examples of raw abstract snippets and corresponding LLM-generated summaries are shown in Tab.~\ref{tab:datasetsumm}. See App.~\ref{app:guided-generation} for a more detailed description of the guidance generation process.

@@

We emphasize that the goal of summarization-via-guided-generation is to increase the signal between text and images by standardizing the captions used for fine-tuning the CLIP model, and compare the quantitative performance of the model vs using the raw abstracts in Sec.~\ref{sec:results}. We also note that, even after summarization, the association signal is expected to be noisy, since aspects of the summarized caption may not be directly descriptive of the observed images, as can be seen from Tab.~\ref{tab:dataset}.

Finally, in order to test a further compression of the proposal abstracts, we use LLM-guided generation to also produce a list of single-concept summaries (e.g., `irregular galaxy', `quasars', `Galactic bulge', \ldots), and test whether this can lead to meaningful, generalizable learned associations with observations. The prompts and schemata for generating these discrete categories and assigning observations to them are described in Apps.~\ref{app:singleconcept} and \ref{app:singleconceptassignments} respectively.

@@

\begin{table}[h!]
      \centering
      \begin{tabular}{m{0.21\textwidth} p{1.9cm} p{1.9cm} m{8cm}}
          \toprule
          \centering \bfseries Image & \centering \bfseries Obs. cycle & \centering \bfseries Prop. ID & \centering \bfseries LLM-extracted summary \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_0.pdf} & \centering \input{\thedatafolder/cycle_0.txt} & \centering \input{\thedatafolder/id_0.txt} &  {\scriptsize \input{\thedatafolder/sum_0.txt}} \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_1.pdf} & \centering \input{\thedatafolder/cycle_1.txt} & \centering \input{\thedatafolder/id_1.txt} &  {\scriptsize \input{\thedatafolder/sum_1.txt}} \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_2.pdf} & \centering \input{\thedatafolder/cycle_2.txt} & \centering \input{\thedatafolder/id_2.txt} &  {\scriptsize \input{\thedatafolder/sum_2.txt}} \tabularnewline
          \midrule
          \centering \includegraphics[width=0.2\textwidth]{\thedatafolder/img_3.pdf} & \centering \input{\thedatafolder/cycle_3.txt} & \centering \input{\thedatafolder/id_3.txt} &  {\scriptsize \input{\thedatafolder/sum_3.txt}} \tabularnewline
          \bottomrule
      \end{tabular}
      \caption{Examples of images and corresponding captions, constructed using the LLM-extracted summaries. The CLIP model is fine-tuned using these text-image associations.}
      \label{tab:dataset}
  \end{table}

\begin{landscape}
      \begin{table}[h!]
          \centering
          \begin{tabular}{m{1.8cm} m{8cm} m{5cm} m{6cm}}
              \toprule
              \bfseries Prop. ID & \centering\arraybackslash \bfseries Proposal abstract & \multicolumn{2}{c}{\bfseries LLM-extracted summary} \tabularnewline
              \cmidrule(r){3-4}
              & & \centering\arraybackslash \bfseries Objects and phenomena & \centering\arraybackslash \bfseries Science use cases \tabularnewline
              \midrule
              \input{\thedatafolder/id1_0.txt} & {\scriptsize \input{\thedatafolder/abs1_0.txt}} & {\scriptsize \input{\thedatafolder/obj1_0.txt}} & {\scriptsize \input{\thedatafolder/sci1_0.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_1.txt} & {\scriptsize \input{\thedatafolder/abs1_1.txt}} & {\scriptsize \input{\thedatafolder/obj1_1.txt}} & {\scriptsize \input{\thedatafolder/sci1_1.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_2.txt} & {\scriptsize \input{\thedatafolder/abs1_2.txt}} & {\scriptsize \input{\thedatafolder/obj1_2.txt}} & {\scriptsize \input{\thedatafolder/sci1_2.txt}} \tabularnewline
              \midrule
              \input{\thedatafolder/id1_3.txt} & {\scriptsize \input{\thedatafolder/abs1_3.txt}} & {\scriptsize \input{\thedatafolder/obj1_3.txt}} & {\scriptsize \input{\thedatafolder/sci1_3.txt}} \tabularnewline
              \bottomrule
          \end{tabular}
          \caption{Examples of the initial parts of raw proposal abstracts (second column) and LLM (\textsc{Mixtral-8x7B})-extracted summaries (rightmost two columns), separately extracting objects and phenomena as well as potential downstream science use cases.}
          \label{tab:datasetsumm}
      \end{table}
  \end{landscape}

@@

Our goal is to learn a semantically  meaningful joint representation between HST observation images and natural language. We leverage the strong generalization capabilities demonstrated by pre-trained CLIP models and adapt these to work with domain-specific \hubble data via fine tuning.

@@

CLIP \citep[Contrastive Language-Image Pretraining;][]{radford2021learning} is a multi-modal model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss. Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$. Image and text encoders $f: I \rightarrow \mathbb R^{n_\text{emb}}$ and $g: T \rightarrow \mathbb R^{n_\text{emb}}$ are used to map images and text to a common embedding space of dimension $n_\text{emb}$. The standard softmax-based bidirectional variation of the InfoNCE~\citep{oord2018representation} contrastive loss function was introduce for training CLIP-style architectures \citep{radford2021learning}
\begin{equation}
  \mathcal{L}(\mathcal{B})=-\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_i \cdot y_j / \tau}}+\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_j \cdot y_i / \tau}}\right)\label{eq:softmax_loss}
\end{equation}
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter. Note that this loss treats the image and text representations symmetrically, ensuring that the two modalities are on the same footing.

We use the CLIP-ViT-B/16 \citep{radford2021learning} variant as the base pre-trained CLIP model. This model uses a 12-layer, 12-head, 768-embedding dimension vision transformer as the image encoder and a 12-layer, 8-head, 512-embedding dimension sequence transformer as the text backbone. The text encoder has a maximum length of 77 tokens and the image encoder a native resolution of $224\times224$. Linear projection layers map the outputs of the image and text encoders to a common embedding space of dimension $n_\text{emb}=512$. In total, the model has 149,620,737 trainable parameters. The model was originally trained on 400 million image-text pairs from internet data.

@@

The base CLIP model is fine-tuned using the dataset described in Sec.~\ref{sec:dataset}, using either the LLM-summarized data, the raw proposal abstracts, or single-concept summaries. When using raw proposal abstracts, random chunks of the text delimited by periods are selected on the fly to fit within the maximum token length of the text encoder. Images are randomly cropped to the native resolution of the image encoder and randomly rotated at each training step. Given the relatively modest size of the fine-tuning set, a batch size $|\mathcal B| = 32$ is used throughout; larger batch sizes were seen to be susceptible to overfitting. We note that the positive and negative image-text association is noisy and imperfect, since multiple images can be associated with the same abstract.

@@

Models were trained over 20,000 steps with 2000 linear warmup steps 
% and cosine decay 
using the AdamW optimizer \citep{DBLP:conf/iclr/LoshchilovH19,DBLP:journals/corr/KingmaB14} with  %peak 
learning rate of $10^{-5}$ and weight decay $10^{-3}$. Training takes approximately 3 hours on 4 Nvidia A100 GPUs.
Models were instantiated and trained using the \package{Jax} \citep{jax2018github} ecosystem.

@@

The model is evaluated by tracking the loss in Eq.~\ref{eq:softmax_loss} as well as the top-$k\%$ retrieval accuracy on the held out validation set over the course of training. The retrieval accuracy is defined as the fraction of associated captions which fall within the top $k\%$ of captions by cosine similarity of the normalized embeddings $x_i \cdot y_j$, averaged over the images in the validation set.

@@

Figure~\ref{fig:retrieval_acc} shows the contrastive loss (left) and the top-10\% retrieval accuracy on the held out validation set over the course of training, for different training configurations considered. The purple lines show the metrics evaluated when training with batches where the image-text associations are randomly shuffled, serving as a baseline. This baseline is seen to do on par with random expectation, unlike the others, validating the presence of a significant association signal between images and text in the dataset.
%
Interestingly, the base pre-trained model performs better than random expectation, with a top-10\% retrieval accuracy of $\sim 15\%$. We therefore compare the qualitative performance of the base model with the fine-tuned model on downstream retrieval tasks.

The model with LLM-guided summarization (yellow lines) is seen to perform on-par with the model using raw abstracts as captions (orange line), despite the stronger association signal in the summarized dataset curated. Fine-tuning a small MLP head over frozen vision and text backbones (green lines), training from scratch (blue lines), and using the single-concept summaries (red lines) show a non-trivial improvement compared to the random baseline as well as base model, but with deteriorated performance compared to fine-tuning with either summarized or raw abstracts.

\begin{figure*}[!h]
  \includegraphics[width=0.95\textwidth]{plots/val_metrics.pdf}
  \caption{The CLIP contrastive loss from Eq.~\ref{eq:softmax_loss} (left) and the top-10\% retrieval accuracy (right) computed on the validation set over the course of training. Shown for the dataset with summarized abstracts (orange), dataset using raw proposal abstracts (yellow), dataset with single-concept summaries (red), only fine-tuning a small MLP head (green), training from scratch (blue), and trained with shuffled image-text pairs (purple).}
  \label{fig:retrieval_acc}
  \end{figure*}

@@

Having aligned the image and text representations, we can query the validation set using natural language and show the most `relevant' images when ranked by cosine similarity. We show these in Figs.~\ref{fig:tti_base} and \ref{fig:tti} for the base model and fine-tuned model respectively using four simple curated queries: `jupiter', `dwarf galaxy', `SN1987A' (a specific, prominent supernova), and `strong lensing'.

While the base model shows some signs of meaningful retrieval (e.g., the image of jupiter in the first row of Fig.~\ref{fig:tti_base}, and images of galaxies in the second row), it is challenging to discern meaningful strong associations between the retrieved images and corresponding query.

The fine-tuned model (Fig.~\ref{fig:tti}), meanwhile, shows strikingly different behavior. 
% For example, it is able to return images with processing and assembly artifacts particular to HST (e.g., the lines through the middle in some images), which typically receive low similarity scores when using the base model. 
Images of Jupiter are returned for the `jupiter' query, with the planet visible in all four images.
The `dwarf galaxy' images correspond to proposals aiming to measure the kinematics of the stellar cores of dwarf galaxies. Supernova SN1987 itself can be seen in the most relevant images for the `SN1987A' query, with the fourth image corresponding to a supernova remnant. Cluster-scale as well as galaxy-scale gravitational lenses are returned by the `strong lensing' query, with lensing patterns visible in the images.
